---
title: "Sobol with an ODE"
author: "Jackson Coldiron and Kaitlin Castor"
editor: visual
format: 
  html:
    code-fold: true
    embed-resources: true
    theme: sandstone
    toc: false
execute:
  freeze: auto 
  warning: false
  message: false
---

## Introduction
Understanding how forests accumulate carbon over time is essential for evaluating their role in climate mitigation. In this exercise, we model forest carbon growth using a piecewise differential equation that shifts from exponential to logistic growth once a canopy closure threshold is reached. We then apply Sobol global sensitivity analysis to explore how uncertainty in model parameters—such as early growth rate, post-closure growth rate, and carrying capacity—affects estimates of maximum carbon stock over 300 years. This approach helps identify which parameters most influence long-term carbon accumulation and how they may respond to environmental change.

## Model Implementation

```{r}
# Load required libraries
library(tidyverse)
library(deSolve)
library(sensitivity)
library(knitr)
library(kableExtra)
library(here)
```

#### Model Run

```{r}
# Bring in the model
source(here("R","forestgrowth.R"))

# Set the parameters
tm <- seq(from = 1, to = 300)
Pinitial <- 10
gps <- list(thres = 50, K = 250, r = 0.01, g = 2)
res <- ode(Pinitial, tm, forest, gps)
colnames(res) <- c("time", "carbon")

```

#### Results with No Uncertainty

```{r}
#| fig-width: 10
#| fig-cap: "__Figure 1: Growth in Forest Size Over 300 Years with No Uncertainty __. Holding the parameters fixed with given values, forest growth in terms of kilograms of Carbon is shown over 300 years. Forest size grows exponetially until it reaches the canopy closure threshold at 50 kg C. Then growth follows a linear model based on a carrying capacity of 250 kg C."

# Graph the results with a line signally carrying capacity
no_uncert <- ggplot(as.data.frame(res), aes(time, carbon)) +
  geom_point(fill = "steelblue") +
  labs(x = "Time (years)",
       y = "Forest Size (kg C)") +
  theme_minimal()

no_uncert
```

## Sensitivity Analysis

```{r}
# Set distributions for parameters
# Assume parameters are normally distributed with a sd of 10% of mean value
np <- 1000

r <- rnorm(mean = 0.01, sd = 0.1 * 0.01, n = np)
g <- rnorm(mean = 2, sd = 0.1 * 2, n = np)
K <- rnorm(mean = 250, sd = 0.1 * 250, n = np)
thres <- rnorm(mean = 50, sd = 0.1 * 50, n = np)

# Input sample parameters into a df
X1 <- cbind.data.frame(r, g, K, thres)

# Sobol requires two sets of sample parameters
# Repeat sampling
r <- rnorm(mean = 0.01, sd = 0.1 * 0.01, n = np)
g <- rnorm(mean = 2, sd = 0.1 * 2, n = np)
K <- rnorm(mean = 250, sd = 0.1 * 250, n = np)
thres <- rnorm(mean = 50, sd = 0.1 * 50, n = np)

# Input sample parameters into a df
X2 <- cbind.data.frame(r, g, K, thres)
```

#### Sobol

```{r}
# Create a sobol object and get sets of parameters for running the model
sens_forest <- sobolSalt(model = NULL, X1, X2, nboot = 300)
colnames(sens_forest$X) <- c("r", "g", "K", "thres")

# Define the metric function
metric <- function(res) {
  max <- max(res$carbon)
  return(list(max = max))
}

# Use a wrapper function to return just the map
p_wrapper <- function(r, g, K, thres, Pinitial, simtimes, func, metric) {
  parms <- list(r = r, K = K, g = g, thres = thres)
  result <- ode(y = Pinitial, times = simtimes, func = func, parms = parms)
  result <- as.data.frame(result)
  colnames(result) <- c("time", "carbon")
  # get metrics
  metrics <- metric(as.data.frame(result))
  return(metrics$max)
}

# Run the model with the parameters
allresults <- as.data.frame(sens_forest$X) |>
  pmap(p_wrapper, Pinitial = Pinitial, simtimes = tm, func = forest, metric = metric)
```

#### Visualize Results

```{r}
#| fig-width: 10
#| fig-cap: "__Figure 2: Maximum Forest Size Distribution.__ Distribution of maximum carbon for 300 years, a proxy for maximum forest size, with uncertainty. Parameters vary for pre canopy closure growth rate, post-canopy closure growth rate, and canopy closure threshold and carrying capacity. Given the uncertainty in our parameters, maximum forest size averages at 182 kg C. The interquartile range is 171 - 192 kg C. The maximum and minimum for maximum forest size are 237 and 121 kg C, respectively."

# Convert vector of results to a data frame for plotting
allres <- data.frame(max_C = unlist(allresults))

#create a boxplot
ggplot(allres, aes(y = max_C)) +
  geom_boxplot(fill = "steelblue") +
  labs(title = "Distribution of Maximum Carbon Stock (C)",
       y = "Max Carbon (kgC)") +
  theme_minimal()

```

#### Sobol Indices

```{r}
#| fig-width: 10
#| fig-cap: "__Table 1: Sobel Indices__ for maximum carbon over 300 years, a proxy for maximum forest size. The First-order index shows the main effect of parameters without co-varience."

sens_forest$y <- unlist(allresults)
sens_forest <- sensitivity::tell(sens_forest, sens_forest$Y)

# first-order index (main effect without co-variance)
rownames(sens_forest$S) <- c("r", "g", "k", "thresh")
#sens_forest$S

kable(sens_forest$S, 
      caption = "first-order index") |> 
  kable_styling(position = "left")
```

```{r}
#| fig-width: 10
#| fig-cap: "__Table 2: Sobel Indices __ for maximum carbon over 300 years, a proxy for maximum forest size. The total sensitivity index shows how the parameters affect the model, accounting for when all parameters interact together."

# total sensitivity index -note that this partitions the output variance
rownames(sens_forest$T) <- c("r", "g", "k", "thresh")
#sens_forest$T

kable(sens_forest$T, 
      caption = "total sensitivity index") |> 
  kable_styling(position = "right")
```

## Discussion

Our Sobol indices quantify how each parameter contributes to the variance observed in modeled maximum forest size over 300 years. The pre-canopy closure growth rate (r) and carrying capacity (K) exhibit the strongest influence, each contributing approximately 35% to both first-order (individual) and total-effect (interaction-inclusive) sensitivity indices. In contrast, the canopy closure threshold (thres) shows the weakest influence, with only 10% first-order and 7% total-effect contribution. This suggests that uncertainty in thres has minimal impact on long-term forest carbon accumulation. These results imply that as climate change alters variability in early growth rates and carrying capacities—through shifts in temperature, water availability, or soil productivity—it may substantially affect maximum forest size. Accurately estimating these parameters will therefore be critical for reliable forest carbon modeling under future climate scenarios.

## Extra Credit
```{r}
# New metric: forest size at year 100
metric_100yr <- function(res) {
  value_100 <- res$carbon[which.min(abs(res$time - 100))]
  return(list(y100 = value_100))
}

# Modify wrapper function
p_wrapper_100yr <- function(r, g, K, thres, Pinitial, simtimes, func) {
  parms <- list(r = r, K = K, g = g, thres = thres)
  result <- ode(y = Pinitial, times = simtimes, func = func, parms = parms)
  result_df <- as.data.frame(result)
  colnames(result_df) <- c("time", "carbon")
  carbon_100 <- result_df$carbon[which.min(abs(result_df$time - 100))]
  return(carbon_100)
}

# Run a new Sobol analysis
sens_100yr <- sobolSalt(model = NULL, X1 = X1, X2 = X2, n = np, nboot = 0)
colnames(sens_100yr$X) <- c("r", "g", "K", "thres")

results_100yr <- as.data.frame(sens_100yr$X) %>%
  pmap_dbl(p_wrapper_100yr, Pinitial = Pinitial, simtimes = tm, func = forest)

sens_100yr$y <- results_100yr
sens_100yr <- sensitivity::tell(sens_100yr, sens_100yr$y)

# Add rownames and extract 'original' values (the index estimates)
s_100 <- sens_100yr$S$original
s_300 <- sens_forest$S$original
t_100 <- sens_100yr$T$original
t_300 <- sens_forest$T$original

# Combine into one comparison data frame
sobol_compare <- data.frame(
  Parameter = c("r", "g", "K", "thresh"),
  `S_100yr` = round(s_100, 3),
  `S_300yr` = round(s_300, 3),
  `T_100yr` = round(t_100, 3),
  `T_300yr` = round(t_300, 3)
)

kable(sobol_compare, caption = "Comparison of Sobol Indices for Forest Size at 100 Years vs Maximum Forest Size (300 Years)") %>%
  kable_styling(full_width = F, position = "center") %>%
  column_spec(1, bold = TRUE)

```
The Sobol analysis reveals a notable shift in parameter influence between short- and long-term forest growth outcomes. At 100 years, pre-canopy closure growth rate (r) overwhelmingly dominates the variance in forest size, with a first-order Sobol index of 1.000, indicating it is the sole driver of variation at that timescale. In contrast, for maximum forest size over 300 years, the influence of r diminishes (S = 0.338), while carrying capacity (K) and post-canopy growth rate (g) become much more influential (S = 0.360 and 0.221, respectively). This suggests that while early growth dynamics are critical in the short term, long-term carbon accumulation depends more on the system’s biophysical limits and slower, density-dependent processes. The canopy closure threshold (thresh) remains a minor contributor in both scenarios.
